---
title: "Interactive Visualization"
author: "EE375 Lecture 07"
output:
  ioslides_presentation:
    incremental: true
---

```{r, echo=FALSE, include=FALSE}
library(plotly)
library(tidyverse)
library(animation)
library(gganimate)
library(shiny)
##library(STRbook)  #devtools::install_github("andrewzm/STRbook")
download.file("https://github.com/andrewzm/STRbook/raw/master/data/NOAA_df_1990.rda","NOAA_df_1990.rda")

data(diamonds,package="ggplot2")

### load met data from previous labs
met = read.csv("met_hourly.csv")
met2 = met[sample.int(nrow(met),500),] ## subsample data

## load NOAA data
#data("NOAA_df_1990", package = "STRbook")
load("NOAA_df_1990.rda")

Tmax <- filter(NOAA_df_1990,     # subset the data
              proc == "Tmax" &   # only max temperature
              month %in% 7:8 &   # May to September
              year == 1993)      # year of 1993
Tmax$t <- Tmax$julian - 728049
```

## Outline

- Why interactive graphics
- Making individual plots interactive
  - Plotly & ggplotly()
- Making movies
  - library(animation)
- Making interactive “apps”
  - Shiny

## Why interactive?
- High-dimensional data
   - e.g. spatio-temporal
- Identifying structure in data
   - interactive querying
- Diagnosing models
- Aid in exploratory analyses
- Useful for decision support tools

# Next up: interactive plots

# Making individual plots interactive

## Plotly
- Plotly.js is a widely used JavaScript library for interactive graphics
- library(plotly) is an R interface
 - plot_ly() direct interface
 - ggplotly() converts ggplot objects
 - plot_mapbox() and plot_geo() for spatial

```{r}
library(plotly)
```

---

```{r}
plot_ly(diamonds) %>% add_histogram(x = ~cut)
```

---

```{r}
plot_ly(diamonds) %>% add_histogram(x = ~cut, color = ~clarity, colors = "Accent")
```

---

```{r}
 plotly::plot_ly(met) %>%
 plotly::add_trace(x = ~AirTemp, y = ~LongWave,type='scatter',mode= 'markers')
```

## ggplotly

```{r}
g =  ggplot(met) + aes(x = AirTemp, y = LongWave) +
 geom_point(colour = "blue") + theme_minimal()
g
```

---

```{r}
ggplotly(g)
```

---

```{r}
g =  ggplot(met) + aes(x = AirTemp, y = LongWave) +
 geom_hex() + theme_minimal()
ggplotly(g)
```

## Additional resources

- Interactive web-based data visualization with R, plotly, and shiny by Carson Sievert
  - https://plotly-r.com/
- htmlwidgets
  - http://gallery.htmlwidgets.org/
  - R libraries that link to other JavaScript interactive libraries
- flexdashboard
  - composes multiple htmlwidgets into a dashboard
- crosstalk
  - linked brushing across htmlwidgets

# Next up: Animation

# Animation

## Idea
- use code to draw a series of data visualizations (e.g. maps through time)
- combine images into movie format (gif, mp4, etc.) with some delay between images

## Example: NOAA met

```{r}
head(Tmax)
```


## Write function to draw plots

```{r}
Tmax_t <- function(tau) {                     ## tau = day of year to plot
    Tmax_sub <- filter(Tmax, t == tau)        # subset data
    ggplot(Tmax_sub) +
        geom_point(aes(x = lon,y = lat, colour = z),   # scatter plot with temperature as color
                   size = 4) +                         # pt. size
        scale_colour_distiller(palette = "Spectral", guide = "colourbar",   # scale
                               name = "z", limits = c(40, 110)) +
        theme_bw() # B&W theme
}
```

## Write function to loop over plots
```{r}
gen_anim <- function() {
  lim_t <- range(Tmax$t)            # time range
   for(t in lim_t[1]:lim_t[2]){  # for each time point
      plot(Tmax_t(t))            # plot data at this time point
   }
}
```

## Animate

```{r}
ani.options(interval = 0.2)     # 0.2s interval between frames
saveHTML(gen_anim(),            # run the main function
        autoplay = FALSE,      # do not play on load
        loop = FALSE,          # do not loop
        verbose = FALSE,       # no verbose
        outdir = ".",          # save to current dir
        single.opts = "'controls': ['first', 'previous',
                                    'play', 'next', 'last',
                                     'loop', 'speed'],
                                     'delayMin': 0",
        htmlfile = "NOAA_anim.html")  # save filename
```


## library(animation)

- **saveGIF**	  - Convert images to a single animation file (typically GIF) using ImageMagick or GraphicsMagick
- **saveHTML**	- Insert animations into an HTML page
- **saveLatex**	- Insert animations into a LaTeX document and compile it
- **saveMovie**	- Convert images to a single animation file (typically GIF) using ImageMagick or GraphicsMagick
- **saveSWF**	  - Convert images to Flash animations
- **saveVideo**	- Convert a sequence of images to a video by FFmpeg

Some of these require additional software (e.g. ImageMagick)

Library comes with a lot of demos

## library(gganimate)

```{r}
a1 <- ggplot(Tmax) + geom_point(aes(x = lon,y = lat, colour = z), size = 4) +   # scatter plot with temperature as color
    scale_colour_distiller(palette = "Spectral", guide = "colourbar", name = "z", limits = c(40, 110)) +
    theme_bw() + # B&W theme
    # gganimate specific bits:
    transition_time(t) # one image per t
anim_save("NOAA.gif",a1)
```

## inline animation

```{r}
animate(a1,fps=5) ## fps = frames per second = 1/interval
```

# Next up: R Shiny
