---
title: "Shiny: R-based interactive web apps"
author: "EE375 Lecture 07"
output:
  ioslides_presentation:
    incremental: true
runtime: shiny
---

## Reactive programming

- Event driven
  - Script no longer run once sequentially
  - Functions re-run when inputs change
- Code a user interface
  - "widgets" control **inputs**
- Code that respond to widgets
  - return **outputs** that update interface
- In addition to webpages, Shiny can be embedded in Rmd
 
## https://shiny.rstudio.com/gallery/

- Gallery of examples (with code)
- Examples of widgets

## App development

- Start with _working_ non-interactive code, default **inputs**
- Wrap outputs in `render` functions
  - DataTable, Image, Plot, CachedPlot, Print, Table, Text, UI
- Create input widget for _one_ thing
- Incrementally add complexity
- Check app functions after each step
  - If things break, easier to debug just the most recent change
 
## Start from working Code

```
n_breaks = 20
bw_adjust = 1
hist(faithful$eruptions, probability = TRUE, breaks = as.numeric(n_breaks),
    xlab = "Duration (minutes)", main = "Geyser eruption duration")
lines(density(faithful$eruptions, adjust = bw_adjust), col = "blue")
```

---

```{r, echo=FALSE}
n_breaks = 20
bw_adjust = 1
hist(faithful$eruptions, probability = TRUE, breaks = as.numeric(n_breaks),
    xlab = "Duration (minutes)", main = "Geyser eruption duration")
lines(density(faithful$eruptions, adjust = bw_adjust), col = "blue")
```

## Wrap outputs

```
shiny::renderPlot({
  n_breaks = 20
  bw_adjust = 1
  hist(faithful$eruptions, probability = TRUE, breaks = as.numeric(n_breaks),
       xlab = "Duration (minutes)", main = "Geyser eruption duration")
  lines(density(faithful$eruptions, adjust = bw_adjust), col = "blue")
})
```

## Add an input widget

```
shiny::inputPanel(
  shiny::selectInput("breaks", label = "Number of bins:",
              choices = c(10, 20, 35, 50), selected = 20)
)
shiny::renderPlot({
  bw_adjust = 1
  hist(faithful$eruptions, probability = TRUE, breaks = as.numeric(input$breaks),
       xlab = "Duration (minutes)", main = "Geyser eruption duration")
  lines(density(faithful$eruptions, adjust = bw_adjust), col = "blue")
})
```

---

```{r,echo=FALSE}
shiny::inputPanel(
  shiny::selectInput("breaks", label = "Number of bins:",
              choices = c(10, 20, 35, 50), selected = 20)
)
shiny::renderPlot({
  bw_adjust = 1
  hist(faithful$eruptions, probability = TRUE, breaks = as.numeric(input$breaks),
       xlab = "Duration (minutes)", main = "Geyser eruption duration")
  lines(density(faithful$eruptions, adjust = bw_adjust), col = "blue")
})
```

## Add complexity incrementally

```
shiny::inputPanel(
  shiny::selectInput("n_breaks", label = "Number of bins:",
              choices = c(10, 20, 35, 50), selected = 20),
  shiny::sliderInput("bw_adjust", label = "Bandwidth adjustment:",
              min = 0.1, max = 2, value = 1, step = 0.1)
)
shiny::renderPlot({
  hist(faithful$eruptions, probability = TRUE, breaks = as.numeric(input$n_breaks),
       xlab = "Duration (minutes)", main = "Geyser eruption duration")
  lines(density(faithful$eruptions, adjust = input$bw_adjust), col = "blue")
})
```

---

```{r,echo=FALSE}
shiny::inputPanel(
  shiny::selectInput("n_breaks", label = "Number of bins:",
              choices = c(10, 20, 35, 50), selected = 20),
  shiny::sliderInput("bw_adjust", label = "Bandwidth adjustment:",
              min = 0.1, max = 2, value = 1, step = 0.1)
)
shiny::renderPlot({
  hist(faithful$eruptions, probability = TRUE, breaks = as.numeric(input$n_breaks),
       xlab = "Duration (minutes)", main = "Geyser eruption duration")
  lines(density(faithful$eruptions, adjust = input$bw_adjust), col = "blue")
})
```
